Typing in tandem: language planning in multi-sentence text production is fundamentally parallel

### 2024-10-26 report/appendix/sim.Rmd
After review update

A possible concern with these results -- substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always lead to a better fit. We addressed this concern before by using cross-validation techniques for model comparison which is preventing overfitting models. To address this concern we repeated a comparison for a single distribution model and a two-distributions mixture model for two sets of simulated data that were simulated either with a single distribution or two-distributional process at heart. In other words, this allows us to test the predictive performance of our models in a context where we know the true underlying data generating process -- single distribution vs two distributions -- and we can test whether these models can successfully uncover the true parameter value.

In particular, the first data set was simulated using a two-distributions mixture model with two mixture components similar to the process described above (equation \ref{eq:bimodcon}). This process and the corresponding Bayesian model is summarised in equation \ref{eq:simmog}.

\begin{equation}
\begin{aligned}
(\#eq:simmog)
\text{y} \sim\text{ } & \theta \times \text{log}\mathcal{N}(\beta + \delta, \sigma^2_1) +\\
& (1 - \theta) \times \text{log}\mathcal{N}(\beta, \sigma^2_2)\\
\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
		& \sigma_{1}^2 > \sigma_{2}^2
\end{aligned}
\end{equation}

This model is largely identical to the models before but reduced to its main parameters (but not mixed effects for participants). The model includes two log-normal distributions with a mixing proportion $\theta$ of which the distribution of shorter values has a mean of $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of longer values is constrained to have a mean that is larger by a factor of $\delta$ and has a larger standard deviation $\sigma^2_1$.

In R, data that follow the model in equation \ref{eq:simmog} can be simulate using a helper function called `molg` (mixture of log Gaussians).

```{r eval = F, echo = T}
# Function for a mixture of log Gaussians
molg <- function(n, theta, mu1, mu2, sigma1, sigma2) {
  y0 <- rlnorm(n, mean = mu1, sd = sigma1)
  y1 <- rlnorm(n, mean = mu2, sd = sigma2)
  mix <- rbinom(n, size = 1, prob = theta)
  y <- y0 * (1 - mix) + y1 * mix 
}

N <- 1000 # number of subjects
beta <- 5 # mean of fluent key intervals (in log)
delta <- 1 # slowdown for hesitant interkey intervals (in log)
theta <- .35 # propotion of hesitations
sigma <- c(.25, .5) # trial-by-trial error

# Generate data from a mixture of two log-Gaussians
y <- molg(n = N, 
          theta = theta,
          mu1 = beta,
          mu2 = beta + delta,
          sigma1 = sigma[1],
          sigma2 = sigma[2])

@@

The second data set was generated with a single distribution log-Gaussian process. The model and its corresponding Bayesian model is summarised in equation \ref{eq:simuv}.

\begin{equation}
\begin{aligned}
(\#eq:simuv)
\text{y} \sim\text{ }& \text{log}\mathcal{N}(\beta, \sigma^2)\\
\text{constraint: } & \sigma^2>0
\end{aligned}
\end{equation}

Data following the distirbution in equation \ref{eq:simuv} can be simulated using a few lines of R code:

```{r echo = T, eval=F}
N <- 1000 # number of subjects
beta <- 5 # Population mean
sigma <- .25 # trial-by-trial error

# Generate data from a log normal distribution
y <- rlnorm(n = N, mean = beta, sd = sigma)
```

Again, this model is a simplified version of the single distribution models used in the main analysis. The model assumes a log-Gaussian distribution with a mean parameter $\beta$ and a standard deviation $\sigma^2$.

The parameter values used for each of the two data simulations can be found in Table \ref{tab:simparam}. The simulated data are visualised in Figure \ref{fig:simdata}. Parameter values were chosen so that the simulated data are roughly similarly distributed to keystroke transitions.

```{r simdata, fig.cap= "Data simulated with a two-distribution (grey) and a single-distribution (yellow) random data generating process. The x-axis showing the outcome y was log-scaled for visability.", fig.height=3.5, fig.width=5}
files <- list.files("../../mixture-model-sim/data", full.names = T)

@@

For each of these two data sets we simulated 1,000 observations. We run 2 models -- a two-distribution mixture model and a single-distribution model -- each for both data sets. Models were run with 3 chains, with each 6,000 iterations of which 3,000 were warmup. Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}. The parameters are shown by type of data generating process along with the true parameter values. Parameter value estimates are shown by type of Bayesian model. The results show that either of the two Bayesian models succesfully uncovered the model parameters of the data with its corresponding data generating process, as shown in Table \ref{tab:simparam}, but less so when the model was applied to data generated with the incorrect underlying process. Particularly the mixing proportion $\theta$ and the slowdown parameter $\delta$ were not uncovered at all by they single-distribution model.

@@

names(table) <- c("Model", "Parameter with true value", "Two-distributions model", "Single-distributions model")

apa_table(table[,-1], 
          escape = FALSE,
      col_spanners = list("Estimate with 95\\% PI" = 2:3),
      caption = "Uncovered parameter estimates with 95\\% probability interval (PI) and true parameter values for each simulated data set and by model and their respective parameters.",
      align = c("l", rep("r", 2)),
      stub_indents = list(`Two-distributions data` = 1:5, 
                          `Single-distribution data` = 6:7))

@@

We used LOO-CV to compare the predictive performance of the two models for each data generating process. The model comparisons can be found for each data generating process in Table \ref{tab:loossim}. The results rule out the possibility that the mixture model does always lead to higher predictive performance. Indeed, the mixture model showed a slightly lower predictive performance for the data that were generated with a single-distribution process. However, for the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance. In fact, the mixture model's predictive performance is 11.6 standard errors higher compared to the single-distribution model. In comparison, for the single-distribution data, the two-distributions mixture model was negligibly different from the single-distribution model with a difference of 0.77 standard errors. Thus, the single distribution is more parsimonious.

@@

apa_table(table[,-1], 
          escape = FALSE,
      caption ="Model comparisons by data set. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
      align =c("l", rep("r", 2)),
      stub_indents = list(`Data: Two-distribution mixture process` = 1:2, 
                          `Data: Single-distribution process` = 3:4))

>>>>

A possible concern with our results -- the substantially better predictive performance for two-distribution mixture models -- is that, in principle, as the mixture model has more parameters it might always outperform single-distribution models. We addressed this concern before by using cross-validation techniques for model comparison which prevent overfit by penalising models with more parameters. To further address this concern we evaluated models -- similar to the ones used in the main text -- for two sets of simulated data. Data were simulated using two random number generators, one that samples data from a single distribution and another that samples data from a weighted combination of two distributions. These simulated datasets allow us to test the predictive performance of our models in a context where we know the true underlying data generating process.

The first dataset was simulated from a weighted mixture of two log-normal distributions similar to the process described above (equation \ref{eq:bimodcon}). This process and the corresponding Bayesian model that we used to parameter estimation is summarised in equation \ref{eq:simmog}.

\begin{equation}
\begin{aligned}
(\#eq:simmog)
\text{y} \sim\text{ } & \theta \times \log\mathcal{N}(\beta + \delta, \sigma^2_1) +\\
& (1 - \theta) \times \log\mathcal{N}(\beta, \sigma^2_2)\\
\text{constraint: } & \delta, \sigma_\text{2}^2, \sigma_\text{1}^2>0\\
		& \sigma_{1}^2 > \sigma_{2}^2
\end{aligned}
\end{equation}

The equivalent Bayesian model is largely identical to the model used in the main text but does not assume different parameters for transition locations and does not include random effects for participants. This is because we neither simulated data for different transition locations (or other factors) or repeated measures for participants. The model assumes two data generating processes with each assuming a log-normal distribution with a mixing proportion $\theta$. The distribution of smaller values has a mean $\beta$ and a standard deviation $\sigma^2_2$; the second distribution of larger values is constrained to have a mean that is $\delta$ units larger than $\beta$ and has a standard deviation $\sigma^2_1$ larger than the distribution with the central tendency $\beta$.

In R we can simulate data for equation \ref{eq:simmog} using the helper function below.

```{r eval = F, echo = T}
# Function for a mixture of two log-Gaussians
molg <- function(n, theta, mu1, mu2, sigma1, sigma2) {
  y0 <- rlnorm(n, mean = mu1, sd = sigma1)
  y1 <- rlnorm(n, mean = mu2, sd = sigma2)
  mix <- rbinom(n, size = 1, prob = theta)
  y <- y0 * (1 - mix) + y1 * mix 
}

N <- 1000 # number of participants
beta <- 5 # mean of fluent interkey intervals (in log)
delta <- 1 # increment for hesitant interkey intervals (in log)
theta <- .35 # proportion of hesitations
sigma <- c(.25, .5) # error variance

# Simulate data from a mixture of two log-Gaussians
y <- molg(n = N, 
          theta = theta,
          mu1 = beta,
          mu2 = beta + delta,
          sigma1 = sigma[1],
          sigma2 = sigma[2])

@@

A second dataset was simulated coming from a single log-normal distribution following equation \ref{eq:simuv}. An equivalent Bayesian model was implemented for parameter estimation.

\begin{equation}
\begin{aligned}
(\#eq:simuv)
\text{y} \sim\text{ }& \log\mathcal{N}(\beta, \sigma^2)\\
\text{constraint: } & \sigma^2>0
\end{aligned}
\end{equation}

Data following the distribution in equation \ref{eq:simuv} can be simulated using the following R code:

```{r echo = T, eval=F}
N <- 1000 # number of participants
beta <- 5 # population mean
sigma <- .25 # error variance

# Simulate data from a single log-normal distribution
y <- rlnorm(n = N, mean = beta, sd = sigma)
```

Again, the Bayesian model corresponding to this process is a simplified version of the single-distribution model used in the main text. The model assumes a log-Gaussian distribution with a mean $\beta$ and a standard deviation $\sigma^2$.

Both simulated datasets are visualised in Figure \ref{fig:simdata}. The parameter values used for each of the two data simulations can be seen in Table \ref{tab:simparam}. Parameter values were chosen so that the simulated data are distributed roughly similar to interkey interval data.

```{r simdata, fig.cap= "Data simulated with a two-distribution (yellow) and a single-distribution (grey) random data-generating process. The x-axis showing the outcome y was log-scaled for visibility.", fig.height=3.5, fig.width=5}
files <- list.files("../../mixture-model-sim/data", full.names = T)

@@

For each of these two datasets we simulated 1,000 observations. We fitted 2 models -- the two-distribution mixture model and the single-distribution model described above -- for each dataset. Models were run with 3 chains, with each 6,000 iterations of which 3,000 iterations were discarded as warmup samples. Estimates with 95% probability intervals are shown in Table \ref{tab:simparam}. True and estimated parameter values are shown for each data generating process and Bayesian model. Both models successfully uncovered the model parameters for the datasets simulated with the corresponding underlying process, but less so when the model was applied to the data generated with the other incorrect underlying process.

@@

names(table) <- c("Model", "Parameter with true value", "Two-distribution data", "Single-distribution data")

apa_table(table[,-1], 
          escape = FALSE,
      col_spanners = list("Estimates with 95\\% PIs" = 2:3),
      caption = "Uncovered parameter estimates with 95\\% probability intervals (PI) and true parameter values for each simulated dataset by model.",
      align = c("l", rep("r", 2)),
      stub_indents = list(`Two-distribution model` = 1:5, 
                          `Single-distribution model` = 6:7))

@@

We used leave-one-out cross-validation to compare the predictive performance of the two models for each data-generating process. Model comparisons can be found in Table \ref{tab:loossim}. For the data generated with a two-distribution mixture process, the mixture model shows a substantially higher predictive performance. In fact, the mixture model's predictive performance is 11.6 standard errors higher compared to the single-distribution model. However, the mixture model showed a slightly lower predictive performance (i.e. a difference of 0.77 standard errors) for the single-process data. Thus, the single distribution model is the more parsimonious choice for the single-distribution data. These results rule out the possibility that mixture models always lead to higher predictive performance.

@@

apa_table(table[,-1], 
          escape = FALSE,
      caption ="Model comparisons by dataset. The top row shows the models with the highest predictive performance for each data generating process. Standard error is shown in parentheses.",
     note = "$\\widehat{elpd}$ = predictive performance indicated as expected log pointwise predictive density; $\\Delta\\widehat{elpd}$ = difference in predictive performance relative to the model with the highest predictive performance in the top row.",
      align =c("l", rep("r", 2)),
      stub_indents = list(`Data: Two-distribution mixture process` = 1:2, 
                          `Data: Single-distribution process` = 3:4))

### 2024-10-26 report/appendix/transloc_constr.Rmd
After review update

apa_table(tmp[,-1],
          align =c("l", rep("r", 4)), 
          longtable = T,
          escape = FALSE,
          row.names = T,
          font_size = "footnotesize",
          col.names = c("Comparison", rep(c("Est. [95\\% PIs]", "BF"), 2)),
          stub_indents = list(`\\textbf{C2L1}` = 1:2, 
                              `\\textbf{CATO (non-dyslexic, unmasked)}`= 3:4,
                              `\\textbf{GE2 (unmasked)}` = 5:6,
                              `\\textbf{LIFT}` = 7:8,
                              `\\textbf{PLanTra}` = 9:10,
                              `\\textbf{SPL2 (L1)}` = 11:12),
          #                              `\\textbf{SPL2 (L1; shift + C)}` = 13:14),
          col_spanners = list(" " = c(1),
                          #    "Fluent transitions" = c(2, 3), 
                              "Hesitation slowdown" = c(2, 3), 
                              "Hesitation probability" = c(4, 5)),
          note = "PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
          caption = "Effect of transition location on interkey intervals. Differences are shown on log scale (for durations) and logit scale for hesitation probability. 95\\% PIs in brackets.")

>>>>

apa_table(tmp[,-1],
          align =c("l", rep("r", 4)), 
          longtable = T,
          escape = FALSE,
          row.names = T,
          font_size = "footnotesize",
          col.names = c("Comparison", rep(c("Est. [95\\% PIs]", "BF"), 2)),
          stub_indents = list(`\\textbf{C2L1}` = 1:2, 
                              `\\textbf{CATO (non-dyslexic, unmasked)}`= 3:4,
                              `\\textbf{GE2 (unmasked)}` = 5:6,
                              `\\textbf{LIFT}` = 7:8,
                              `\\textbf{PLanTra}` = 9:10,
                              `\\textbf{SPL2 (L1)}` = 11:12),
          #                              `\\textbf{SPL2 (L1; shift + C)}` = 13:14),
          col_spanners = list("Hesitation duration" = 2:3, 
                              "Hesitation probability" = 4:5),
          note = "PI = probability intervals. BF = evidence in favour of the alternative hypothesis over the null hypothesis.",
          caption = "Transition-location effect. Differences are shown for the hesitation duration (log scale) and the hesitation probability (logit scale); 95\\% PIs in brackets.")

### 2024-10-26 report/manuscript.Rmd
After review update

---
title             : "Typing in tandem: finite mixture models reveal fundamental parallelism in multi-sentence text composition"
shorttitle        : "Typing in tandem"

@@

author: 
  - name          : "Jens Roeser"
    affiliation   : "1"
    address       : "50 Shakespeare St, Nottingham NG1 4FQ"
    corresponding : yes 
    email         : "jens.roeser@ntu.ac.uk"
  - name          : "Rianne Conijn"
    affiliation   : "2"
  - name          : "Evgeny Chukharev"
    affiliation   : "3"
  - name          : "Gunn Helen Ofstad"
    affiliation   : "4"
  - name          : "Mark Torrance"
    affiliation   : "1"

@@

# We are grateful for all authors who made their available (to us), in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues. This work was supported by the National Science Foundation under Grant No. 2016868: "ProWrite: Biometric feedback for improving college students' writing processes." and UKRI ESRC under Grant No. ES/W011832/1: "Can you use it in a sentence?: Establishing how word-production difficulties shape text formation."
# 
# Data and analysis scripts written in R and Stan can be found on OSF (https://osf.io/z65dw/; Roeser et al., 2024). For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see https://rpubs.com/jensroes/mixture-models-tutorial.

@@

keywords: "Cascade; parallel models; serial models; finite mixture models; Bayesian models; text composition; keystrokes"

bibliography      : ["references.bib"]

@@

beta_label <- "Fluent interkey\nintervals (in msecs)"
delta_label <- "Slowdown for hesitant\ninterkey intervals (in secs)"
theta_label <- "Probability of hesitant\ninterkey intervals"
```

@@

@@

```{r child = 'appendix/transloc.Rmd', eval=F}
```

```{r child = 'appendix/shift.Rmd', eval = F}
```

>>>>

---
title             : "Typing in tandem: language planning in multi-sentence text production is fundamentally parallel"
#shorttitle        : "Typing in tandem"
shorttitle        : "Revision of XGE-2024-2060 as invited by the action editor, Michele Diaz, Ph.D."

@@

author: 
  - name          : "Jens Roeser"
    affiliation   : "1"
    address       : "50 Shakespeare St, Nottingham NG1 4FQ"
    corresponding : yes 
    email         : "jens.roeser@ntu.ac.uk"
    role          : 
      - Conceptualisation
      - Data curation
      - Formal analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Project administration
      - Software
      - Visualisation    
      - Writing -- original draft
      - Writing -- review \& editing
  - name          : "Rianne Conijn"
    affiliation   : "2"
    role          :
      - Conceptualisation
      - Investigation
      - Writing -- original draft
      - Writing -- review \& editing
  - name          : "Evgeny Chukharev"
    affiliation   : "3"
    role          :
      - Funding acquisition
      - Investigation
      - Resources  
      - Project administration
      - Software
      - Supervision
  - name          : "Gunn Helen Ofstad"
    affiliation   : "4"
    role          :
      - Investigation
      - Resources

  - name          : "Mark Torrance"
    affiliation   : "1"
    role          :
      - Conceptualisation
      - Data curation
      - Formal analysis
      - Funding acquisition
      - Investigation
      - Methodology
      - Resources  
      - Supervision
      - Visualisation    
      - Writing -- original draft
      - Writing -- review \& editing

@@

authornote: |
  We are grateful for all authors who made their data available to us, in particular Nina Vandermeulen, Alessandra Rossetti, and colleagues, and Vibeke Rønneberg, Per Henning Uppstad, and colleagues at the Norwegian Reading National Centre, University of Stavanger. This work was supported by the National Science Foundation under Grant No. 2016868: "ProWrite: Biometric feedback for improving college students’ writing processes." and UKRI ESRC under Grant No. ES/W011832/1: "Can you use it in a sentence?: Establishing how word-production difficulties shape text formation."

  Data and analysis scripts written in R and Stan can be found on OSF (https://osf.io/z65dw/; Roeser et al., 2024). For a tutorial on Bayesian mixture-model analysis in the context of keystroke data see https://rpubs.com/jensroes/mixture-models-tutorial.

  Parts of this work have been shared through conference presentations, namely at the SIG writing conference in Paris, France (2024) and in Umea, Sweden (2022), and at EARLI SIG Online Measurements in Southampton, UK (2022).

@@

keywords: "Parallel processing; writing; mixture models; language production; keystroke logging"

bibliography      : ["references.bib"]
nocite: | 
  @brysbaert2021improving, @dickey1970weighted, @wagenmakers2010bayesian, @jeffreys1961theory, @lee2014bayesian, @dienes2014using

@@

beta_label <- "Fluent interkey\nintervals"
delta_label <- "Hesitant\ninterkey intervals"
theta_label <- "Probability of hesitant\ninterkey intervals"
```

@@

```{r child = 'appendix/parallel.Rmd'}
```

@@
